Free Probability Analysis for Reflective Surface aided Massive MIMO systems

### 2020-09-19 documentation/Paper/abstract.tex
updates to report

Surfaces with reflective phase control (IRS) have been the subject of interest in recent years as means to better control the wireless propagation environment. This work investigates the impact of an IRS on the capacity of MIMO communication systems. Using the tools of free probability, a system is analyzed in the asymptotic domain to derive an expression for the system capacity. A key finding is that in the asymptotic domain, the choice for the phase of reflected waves has no impact for the asymptotic analysis. Motivated the contradiction between this finding and the substantial previous literature investigating the optimization of the IRS, we consider the convergence of a realistic system towards asymptotic behavior. We then investigate new methods for phase optimization for systems without full channel state information at the IRS.

>>>>

Surfaces which can control the phase of reflected radio waves, known as intelligent reflective surfaces (IRS), have been the subject of interest in recent years as a means to better control the wireless propagation environment. This work investigates the impact of an IRS on the MIMO communication systems. Using the tools of free probability, a system is analyzed in the asymptotic domain to derive an expression for the system capacity. A key finding is that the phase control offered by the IRS has no impact in the asymptotic domain. Motivated by the contradiction between this finding and the substantial previous literature investigating the optimization of the IRS, we consider the convergence of a realistic system towards asymptotic behavior. We then propose an optimization method for the non-convex problem of minimizing the Mean Square Error with respect to the IRS phases.

### 2020-09-19 documentation/Paper/free_probability.tex
updates to report

\begin{itemize}

\item 
	A brief introduction to free probability.
	\par 
	Free probability is the counterpart to classical probability used to analyze cases in which random variables (in this case random matrices) do 
	not commute. In the same way that classical probability provides methods to find the pdf for the sums and products of commuting random
	 variables, free probability provides tools to find the pdfs for products and sums of non-commuting random variables.
	In the following work, a number of tools from free probability will be used. These tools are now introduced with appropriate context and intuition 
	for this investigation but for a thorough coverage of these topics, see \cite{tulino2004random} and \cite{mingo2017free}.
	 \begin{itemize}
	 \item 
	 	The cornerstone of free probability is the concept freeness of two random variables in the context of a projection. In our case, freeness between two 
	 	matrices is given by

>>>>

Equation \eqref{no_csi_capacity_aed} illustrates that the eigenvalue distribution of the received signal covariance matrix is sufficient information to find the capacity of a MIMO channel. For general, random channels, finding this eigenvalue distribution is not a straightforward task.
We now introduce some of the key tools that will be used in order to perform this analysis.
\begin{itemize}
\item 
	A brief introduction to free probability.
	\par 
	Free probability is the counterpart to classical probability used to analyze cases in which random variables (in this case random matrices) do 
	not commute. In the same way that classical probability provides methods to find the pdf for the sums and products of commuting random
	 variables, free probability provides tools to find the pdfs for products and sums of non-commuting random variables.
	In the following work, a number of tools from free probability will be used. These tools are now introduced with appropriate context and intuition 
	for this investigation but for a thorough coverage of these topics, see \cite{tulino2004random} and \cite{mingo2017free}.
	 \begin{itemize}
	 \item 
	 	The cornerstone of free probability is the concept freeness of two random variables in the context of a projection. In our case, freeness between two 
	 	matrices is given by

### 2020-09-19 documentation/Paper/information_theory_background.tex
updates to report

Consider the general MIMO communication system given by 
	\begin{equation}\label{com_system}
	\vy = \mH_{Total}\vx + \vn
	\end{equation}
	in which $\mH_{Total}$ is a random channel. Using methods such as OFDM to transform any multi-path channel into a flat-fading channel, this representation is sufficient to represent any linear channel.
A key metric of interest for this system is the mutual information between the transmitted and received signal given by 
\begin{equation}\label{mut_ent}
I(\vy;\vx) = h(\vy) - h(\vy|\vx).
\end{equation}
If both $\vx$ and $\vn$ are circularly symmetric Gaussian random vectors, $\vy$ is also circularly symmetric Gaussian with entropy given by \cite{telatar1999capacity} 
\begin{equation}\label{entropy}
h(\vy) = \Expect[\Log(|\pi e \mathbf{Q}_{y}|)]
\end{equation}
 for $\mathbf{Q}_{y} = \Expect{[\vy \vy^H]}$. Note also that for a given $\mathbf{Q}_{y}$ the entropy is maximized if and only if $\vy$ is circularly symmetric Gaussian. (explain what implication this has)
For a given transmit covariance matrix, $\mathbf{Q_x}$ and $\Expect\left[\vn\vn^H \right] = \mI$, substituting equation \eqref{entropy} into equation
\eqref{mut_ent} and simplifying gives
\begin{equation}\label{mut_inf}
\Expect \left[\Log(|\mathbf{I} + \mHt \mQ_x \mHt^H|)\right].
\end{equation}
Maximizing the mutual informationwith respect to both the
transmit covariance matrix and phases at the RIS gives an expression for the system capacity

\begin{equation}\label{capacity}
\mathbb{C} = \underset{\boldsymbol{\Phi},\mathbf{Q_x}}{\mathop{max}} \Expect \left[\Log(|\mathbf{I} + \mHt \mQ_x \mHt^H|)\right]
\end{equation}
or equivalently,
\begin{equation}\label{capacity_tricky}
\mathbb{C} = \underset{\boldsymbol{\Phi},\mathbf{Q_x}}{\mathop{max}} \Expect \left[\sum_{i=1}^{N}\Log(1 + \lambda_i)\right].
\end{equation}
where $\lambda_i$ are the eigenvalues of $\mHt \mQ_x \mHt^H$.

Recalling that in general we are interested in evaluating the capacity in order to determine a feasible transmission rate for a given communication channel, equation \eqref{capacity_tricky} poses a couple of hurdles to evaluation. First, we need to be able to evaluate an expectation over the joint pdf $p(\lambda_1 \cdots	 \lambda_N)$. Second, there is an optimization problem whose solution may in general depend on the distribution of the channel. Including this solution in the capacity expression without a closed form may be difficult.
To resolve these hurdles, two simplifications will be used.
First, we will enforce that the transmitter chooses $\mQ_x = \frac{P_{\text{total}}}{N_t}\mathbf{I}$. 
By enforcing a deterministic choice of $\mQ_x$, we no longer have to consider this component of the expectation.
Second we will normalize the channel, and therefore the received power, to allow the received signal covariance matrix eigenvalue distribution to converge. 
These assumptions lead to the form of equation \eqref{capacity}
\begin{equation}\label{no_csi_capacity}
\mathbb{C} = \underset{\phase}{\mathop{max}} \; N_R \Expect \left[\Log(1 + \frac{P_{\text{total}}}{\sigma_nN_t}\lambda_i)\right].
\end{equation}
where $\lambda_i$ are the eigenvalues of $\mHt \mHt^H$.
In this form, we can clearly see the linear scaling of capacity with $N_R$ that has motivated MIMO research.
If we assume that $\Expect\left[ \vy \vy^H \right]$ has a converging eigenvalue distribution as the system dimensions increase to infinity, we can expand the expectation over the AED to get the expression
\begin{equation}\label{no_csi_capacity_aed}
\mathbb{C} = \underset{\phase}{\mathop{max}} \; N_R  \int_{0}^{\infty}\Log(1 + \frac{1}{\sigma_n}x)p_{\lambda\lambda^H}(x) dx
\end{equation}
in which $p_{\lambda\lambda^H}(x)$ is the channel AED and the bounds of the integral reflect the range for the eigenvalues of a positive semidefinite matrix.
Note that because the choice for the distributions of $\vx$ and $\vn$ maximize $h(\vy)$, equation \eqref{capacity} provides an upper bound on the capacity of the channel for any choices of the distribution of $\vx $ with Gaussian noise $\vn$. (Perhaps mention that normally we cannot generate continuous Gaussian x so we use channel coding).
\par
Discuss if $\mQ_x = \frac{P_{\text{total}}}{N_t}\mathbf{I}$ is optimal at high SNR (are we in the low SNR regeime?)
\par
Note that because $\Log$ is a concave function, we can upper bound the mutual information with
\begin{equation}\label{mut_inf}
\Expect[\Log(1 + \frac{1}{\sigma_n}\lambda)] \leq \Log(1 + \frac{1}{\sigma_n}\Expect[\lambda])
\end{equation}

>>>>

Consider the general MIMO communication system given by 
	\begin{equation}\label{com_system}
	\vy = \mH_{Total}\vx + \vn
	\end{equation}
	in which $\mH_{Total}$ is a random channel. Using methods such as OFDM to transform any multi-path channel into a flat-fading channel, this representation is sufficient to represent any linear channel.
A key metric of interest for this system is the mutual information between the transmitted and received signal given by 
\begin{equation}\label{mut_ent}
I(\vy;\vx) = h(\vy) - h(\vy|\vx).
\end{equation}
If both $\vx$ and $\vn$ are circularly symmetric, Gaussian random vectors, $\vy$ is also circularly symmetric and Gaussian with entropy given by \cite{telatar1999capacity} 
\begin{equation}\label{entropy}
h(\vy) = \Expect[\Log(|\pi e \mathbf{Q}_{y}|)]
\end{equation}
 in which $\mathbf{Q}_{y} = \Expect{[\vy \vy^H]}$. Note also that for a given $\mathbf{Q}_{y}$ the entropy is maximized if and only if $\vy$ is circularly symmetric Gaussian. (explain what implication this has)
For a given transmit covariance matrix, $\mathbf{Q_x}$ and $\Expect\left[\vn\vn^H \right] = \mI$ (allow for this to be different), substituting equation \eqref{entropy} into equation
\eqref{mut_ent} and simplifying gives
\begin{equation}\label{mut_inf}
\Expect \left[\Log(|\mathbf{I} + \mHt \mQ_x \mHt^H|)\right].
\end{equation}
Maximizing the mutual information with respect to both the
transmit covariance matrix and phases at the RIS gives an expression for the system capacity

\begin{equation}\label{capacity}
\mathbb{C} = \underset{\boldsymbol{\Phi},\mathbf{Q_x}}{\mathop{max}} \Expect \left[\Log(|\mathbf{I} + \mHt \mQ_x \mHt^H|)\right]
\end{equation}
or equivalently,
\begin{equation}\label{capacity_tricky}
\mathbb{C} = \underset{\boldsymbol{\Phi},\mathbf{Q_x}}{\mathop{max}} \Expect \left[\sum_{i=1}^{N}\Log(1 + \lambda_i)\right].
\end{equation}
in which $\lambda_i$ are the eigenvalues of $\mHt \mQ_x \mHt^H$.

Recalling that in general we are interested in evaluating the capacity in order to determine a feasible transmission rate for a given communication channel, equation \eqref{capacity_tricky} poses a couple of hurdles to evaluation. First, we need to be able to evaluate an expectation over the joint pdf $p(\lambda_1 \cdots	 \lambda_N)$. Second, there is an optimization problem whose solution will generally depend on the distribution of the channel. Including this solution in the capacity expression without a closed form may be difficult.
To resolve these hurdles, two simplifications will be used.
First, we will enforce that the transmitter chooses $\mQ_x = \frac{P_{\text{total}}}{N_t}\mathbf{I}$ (Justify this by saying we assume sufficient SNR?). 
By enforcing a deterministic choice of $\mQ_x$, we no longer have to consider this component of the expectation.
Second we will normalize the channel, and therefore the received power, to allow the received signal covariance matrix eigenvalue distribution to converge. 
With thesis assumptions, equation \eqref{capacity} becomes
\begin{equation}\label{no_csi_capacity}
\mathbb{C} = \underset{\phase}{\mathop{max}} \; N_R \Expect \left[\Log(1 + \frac{P_{\text{total}}}{\sigma_nN_t}\lambda_i)\right].
\end{equation}
This form is useful in observing the linear scaling of capacity with $N_R$ that has motivated MIMO research.
If we assume that $\Expect\left[ \vy \vy^H \right]$ has a converging eigenvalue distribution as the system dimensions increase to infinity, we can expand the expectation over the AED to get the expression
\begin{equation}\label{no_csi_capacity_aed}
\mathbb{C} = \underset{\phase}{\mathop{max}} \; N_R  \int_{0}^{\infty}\Log(1 + \frac{1}{\sigma_n}x)p_{\lambda\lambda^H}(x) dx
\end{equation}
in which $p_{\lambda\lambda^H}(x)$ is the channel AED and the bounds of the integral reflect the range of eigenvalues for a positive semidefinite matrix.
Note that because the choice for the distributions of $\vx$ and $\vn$ maximize $h(\vy)$, equation \eqref{no_csi_capacity_aed} provides an upper bound on the capacity of the channel for any choices of the distribution of $\vx $ with Gaussian noise $\vn$. 
\par
Also note that because $\Log$ is a concave function, we can upper bound the mutual information with
\begin{equation}\label{mut_inf}
\Expect[\Log(1 + \frac{1}{\sigma_n}\lambda)] \leq \Log(1 + \frac{1}{\sigma_n}\Expect[\lambda])
\end{equation}

### 2020-09-19 documentation/Paper/irs_channel.tex
updates to report

Using the approach from Section \ref{Free_Prob_Intro} to find the asymptotic eigenvalue distribution, we now analyze a channel model that is representative of a MIMO communication system with an RIS. 
\par
In general, the received signal from any line of sight path between a transmitter and receiver would contain  sufficient energy to disregard the contribution of any reflected signal (although this component may be very low rank).  Therefore, we focus on a channel model in which all signal power to the receiver is through the path of the RIS  given by

@@

The random channels $\mathbf{H}_{1}$ and $\mathbf{H}_{2}$ represent the random channel gains from transmitter to RIS and from RIS to receiver, respectively. Similarly, $ \mathbf{R}_{T},  \mathbf{R}_{S} $ and  $\mathbf{R}_{S}$ represent the correlation introduced at the transmitter, RIS and receiver, respectively. 

@@

One such model is exponential correlation. Defined by a single parameter $\rho$, the elements of the corresponding correlation matrix are given by $\rho^{i-j}$.

@@

To begin using the tools of free probability, we must first separate out the free sub-components of the covariance matrix given by 

@@

If all system dimensions are $N$, these S-Trasforms (and AEDs) identical so for this section we will assume that all system dimensions are $N$.
Repeated iteration of the rotational property above gives the equivalent matrix

@@

In \cite[Section 4.3.2]{muller2013applications} it is shown that the hermetian matrix $\mH\mH_^H$ with $h_{i,j} \in \NC\{0, \frac{1}{N}\}$
is free 

@@

Without further analysis, this expression shows the interesting result that under the assumptions of Free Probability and equal power transmission, the phases at the RIS (or indeed any unitary matrix in the channel) cancel out of the expression for the AED. For the remaining terms, we simply need an expression for the S-Transform in order to evaluate the asymptotic system capacity.s

@@

a similar cancellation of phases will occur for the RIS channel with an LOS (or non-zero mean) component of the signal.
Note that if we are simply adding a non-zero mean component. We can find the corresponding AED by looking considering the channel without the non-zero mean and then adding the single eigenvalue from the non-zero mean component; in which case the phases will again cancel out.

>>>>

Using the approach from Section \ref{Free_Prob_Intro} to find the asymptotic eigenvalue distribution, we now analyze a channel model that is representative of a MIMO communication system with an IRS. 
\par
In general, the received signal from any line of sight path between a transmitter and receiver would contain  sufficient energy to disregard the contribution of any reflected signal (although this component may be very low rank).  Therefore, we focus on a channel model in which all signal power to the receiver is through the path of the IRS  given by

@@

The random channels $\mathbf{H}_{1}$ and $\mathbf{H}_{2}$ represent the random channel gains from transmitter to IRS and from IRS to receiver, respectively. Similarly, $ \mathbf{R}_{T},  \mathbf{R}_{S} $ and  $\mathbf{R}_{S}$ represent the correlation introduced at the transmitter, IRS and receiver, respectively. 

@@

One commonly used model exponential correlation. Defined by a single parameter $\rho$, the elements of the corresponding correlation matrix are given by $\rho^{i-j}$.

@@

To begin using the tools of free probability, we first separate out the free sub-components of the total covariance matrix given by 

@@

If all system dimensions are $N$, these S-Trasforms equal. For simplified analysis and notation, we will assume that all system dimensions are $N$.
Repeated iteration of the property above gives the equivalent matrix

@@

In \cite[Section 4.3.2]{muller2013applications} it is shown that the hermetian matrix $\mH\mH^H$ with $h_{i,j} \in \CNO$
is free (relative to what?)...

@@

Without further analysis, this expression shows the interesting result that under the assumptions of Free Probability and equal power transmission, the phases at the IRS (or any unitary matrix in the channel) cancel out of the expression for the AED. For the remaining terms, we simply need an expression for the S-Transform in order to evaluate the asymptotic system capacity.s

@@

a similar cancellation of phases will occur for the IRS channel with an LOS (or non-zero mean) component of the signal.
Note that if we are simply adding a non-zero mean component. We can find the corresponding AED by considering the channel without the non-zero mean and then adding the single eigenvalue from the non-zero mean component; in which case the phases will again cancel out.

### 2020-09-19 documentation/Paper/non_asymptotic.tex
updates to report

Up to this point we have shown an analysis which seems to contradict the observations from previous investigations review in Section \ref{Background} showing the benefits of exploiting RIS phases. The key to the coexistence of these results is determining when the asymptotic assumptions required by Free Probability hold. 
As mentioned in in Section \ref{Background}, \cite {} have shown that for matrices of random elements, the  convergence of the matrix behavior towards that predicted by Free Probability is slowed by adding correlations between elements. We would now like to at least numerically investigate this question to investigate the performance of RIS MIMO systems with parameters reflecting state of the art.

 As seen in \ref{Background} current large systems have been shown for 
sizes S= 224. We now consider whether this size is reasonable under correlations realized in systems (NEED TO FIND and justify).

>>>>

Up to this point we have shown an analysis which seems to contradict the observations from previous investigations reviewed in Section \ref{Background} showing the benefits of exploiting phase control at the IRS. The key to the coexistence of these results is determining when the asymptotic assumptions required by Free Probability hold. 
As mentioned in in Section \ref{Background}, \cite {} have shown that for matrices of random elements, the  convergence of the matrix characteristics towards that predicted by Free Probability is slowed by adding correlations between elements. We would now like to at least numerically investigate this question to investigate the performance of RIS MIMO systems with parameters reflecting state of the art.
(This is where being able to analyze the capacity using the AED from above would provide nice results)
 As seen in \ref{Background} current large systems have been shown for 
sizes S= 224. We now consider whether this size is reasonable under correlations realized in systems (NEED TO FIND and justify).

### 2020-09-19 documentation/Paper/paper.tex
Adding SINR/achievable with corrected model variance
updates to report

\title{Free Probability Analysis for Reflect Surface aided Massive MIMO systems}
\author{Peter Hartig}

@@

\chapter{Analyzing the RIS System}\label{system_model}
\section{Analysis Preliminaries}
\subsection{Power Normalization}
\input{power_normalization}
\subsection{Mutual Information}\label{sectiond:mut_info}
\input{information_theory_background}
\subsection{Free Probability}\label{Free_Prob_Intro}
\input{free_probability}
\section{RIS Analysis}
\subsection{The Asyptotic IRS Channel}
\input{irs_channel}
\subsection{Non-Asymptotic Results}
\input{non_asymptotic}
\chapter{Results}\label{Results}
\section{Asymptotic Results}

\subsection{Asymptotic Results Discussion}

>>>>

\title{Free Probability Analysis for Reflective Surface aided Massive MIMO systems}
\author{Peter Hartig}

@@

\chapter{Analyzing the IRS System}\label{system_model}
\section{Analysis Preliminaries}
\subsection{Power Normalization}
\input{power_normalization}
\subsection{Mutual Information}\label{sectiond:mut_info}
\input{information_theory_background}
\subsection{Free Probability}\label{Free_Prob_Intro}
\input{free_probability}
\section{Applying Free Probability to the IRS Chanel}
\subsection{The Asyptotic IRS Channel}
\input{irs_channel}
\subsection{Non-Asymptotic Results}
\input{non_asymptotic}
\chapter{Results}\label{Results}
\section{Asymptotic Results}
\subsection{Asymptotic Results Discussion}

### 2020-09-19 documentation/Paper/power_normalization.tex
updates to report

Consider the SISO channel with an RIS given by $y = \vh_{rs}^H \phase \vh_{st}$. In this case, $ \phase$ should be chosen optimally such that 
the phases all add constructively at the receiver (\ie $h_{rs,i} \phi_k h_{st,j}$ add constructively at the receiver). 
In the case of an i.i.d Rayleigh channel with elements $\CNO$ and $\sigma^2_{\text{noise}} =1$, the SNR at the receiver is given by $\gamma = N$.
Intuitively if the channel gains are normalized such that the received power is normalized for any system size $N$ the IRS offers no gain. 
In this case we can achieve this normalization by enforcing that elements of the a channel matrix $\mH \in \CN^{N \times K}$ have variance $\frac{1}{N}$.
This observation highlights an point that must be considered when performing asymptotic analysis; the received power must be normalized.

>>>>

Consider the SISO channel with an IRS given by $y = \vh_{rs}^H \phase \vh_{st}$. To maximize capacity in this case, $ \phase$ should be chosen such that 
the each incoming signal at the receiver adds constructively (\eg $\angle h_{rs,i} \phi_k h_{st,j} = 0$). 
In the case of an i.i.d Rayleigh channel with elements $\mathcal{CN} (0,1)$ and $\sigma^2_{\text{noise}} =1$, the SNR at the receiver is given by $\gamma = N$.
This observation highlights the point that when performing asymptotic analysis (\ie when $N \rightarrow \infty$) the received power (or metric of interest) should be normalized with respect to the system size.
Extending this to the case of a MIMO channel, for an $N_R \times N_T$ channel $\mH$, we can achieve this normalization by enforcing that elements of the a channel matrix  have variance $\frac{1}{N_R}$ such that even for concatenated MIMO channels (\eg $\mH_2 \mH_1$) the power of the received signal is still
 $\Expect \left[ \trace \left(\mH_2 \mH_1\vx (\mH_2 \mH_1\vx)^H \right)\right] = 1$ when the total transmit power is normalized to 1. 
 (Discuss why this is important to consider the conservation of energy?
